Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-24 text/back_end_KS.qmd
Dallas rq2

The six Geomarkets within the Dallas-Fort Worth metropolitan area are illustrated in @fig-dallas-maps by total population across the three census periods. The Dallas-Fort Worth region represents a unique metroplex case formed by the convergence of two separate metropolitan areas, which creates a unique urban spatial structure that differs from the classic model originally proposed by @RN4982. As depicted in @fig-dallas-maps, two distinct inner-city core Geomarkets (City of Dallas and City of Fort Worth) are seperated by a centrally located suburban Geomarket encompassing Irving, Arlington, & Grand Prairie. These central Geomarkets are surrounded by the more conventional outre ring of suburban Geomarkets (Dallas County, Collin and Rockwall Counties, and West of Dallas/Ft. Worth Metroplex).

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased substantially from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps showcases all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Ft. Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers - City of Forth Worth and City of Dallas - into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket.

Figure X presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Ft. Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian Pacific Islander), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian Pacific Islander). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian Pacific Islander residents.

These shifts in racial and ethnic composition within and across Geomarkets became more pronounced between 2000 and 2020. The Collin and Rockwall and West of Dallas/Fort Worth Metroplex Geomarkets experienced continued declines in their share of White residents. However, both Geomarkets remained predominantly White (58% and 63%, respectively) relative to all other Geomarkets. In contrast, all other Geomarkets experienced continued increases in the shares of Black and Hispanic residents, resulting in a diminished previously White-majority populations for Dallas CO (33% White, 23% Black, 32% Hispanic; 9% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (30% White, 20% Black, 36% Hispanic; 12% Asian Pacific Islander), and City of Fort Worth (46% White, 15% Black, 33% Hispanic; 4% Asian Pacific Islander). Lastly, continuous increases in the shares of Black and Hispanic residents further diversified the City of Dallas, making it the most racially/ethnically diverse Geomarket in the Dallas-Fort Worth metropolitan area.

Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in Figure X for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Ft. Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remain relatively stable and continued to rank at the lower levels of affluence within the metropolitan area. \[KS: NEED TO CHECK THIS; I THINK CITY OF DALLAS AND CITY OF FORT WORTH ARE MIXED UP BETWEEN GRAPH VERSUS TABLE\]

Figure X also illustrates changes in the percentage of residents living in poverty within Geomarkets over time, revealing shifting dynamics across Geomarkets similar to those observed in median household income. In 1980, poverty rates were comparably low in both for Collin & Rockwall Counties and Dallas County, moderate in the Irving, Arlington, and Grand Prarie and West of Dallas/Ft. Worth Metroplex Geomarkets, and higherst for City of Dallas and City of Fort Worth Geomarkets. Overtime, poverty rates declined in Collin & Rockwall Counties and West of Dallas/Ft. Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in a widening gap in poverty levels across the Dallas-Fort Worth metropolitan area. A similar widening disparity is evident evident in educational attainment. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the Collin & Rockwall Counties Geomarket experienced a substantially greater increase - from 23% in 1980 to 50% in 2020- maintaining the gap with other Geomarkets in terms of educational attainment.

>>>>

The six Geomarkets within the Dallas-Fort Worth metropolitan area are illustrated in @fig-dallas-maps by total population across the three census periods. The Dallas-Fort Worth region represents a unique metroplex case formed by the convergence of two separate metropolitan areas, which creates a unique urban spatial structure that differs from the classic model originally proposed by @RN4982. As depicted in @fig-dallas-maps, two distinct inner-city core Geomarkets (City of Dallas and City of Fort Worth) are seperated by a centrally located suburban Geomarket encompassing Irving, Arlington, & Grand Prairie. These central Geomarkets are surrounded by the more conventional outer ring of suburban Geomarkets (Dallas County, Collin and Rockwall Counties, and West of Dallas/Ft. Worth Metroplex).

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased substantially from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps showcases all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers - City of Forth Worth and City of Dallas - into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket.

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Ft. Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian Pacific Islander), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian Pacific Islander). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian Pacific Islander residents.

These shifts in racial and ethnic composition within and across Geomarkets became more pronounced between 2000 and 2020. The Collin and Rockwall and West of Dallas/Fort Worth Metroplex Geomarkets experienced continued declines in their share of White residents. However, both Geomarkets remained predominantly White (58% and 63%, respectively). In contrast, all other Geomarkets experienced continued increases in the shares of Black and Hispanic residents, resulting in a diminished previously White-majority populations for Dallas County (33% White, 23% Black, 32% Hispanic; 9% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (30% White, 20% Black, 36% Hispanic; 12% Asian Pacific Islander), and City of Fort Worth (46% White, 15% Black, 33% Hispanic; 4% Asian Pacific Islander). Lastly, continuous increases in the shares of Black and Hispanic residents further diversified the City of Dallas, making it the most racially/ethnically diverse Geomarket in the Dallas-Fort Worth metropolitan area.

Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in @fig-dallas-rq1-ses for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remained relatively stable and continued to rank at the lower levels of affluence within the metropolitan area. \[KS: NEED TO CHECK THIS; I THINK CITY OF DALLAS AND CITY OF FORT WORTH ARE MIXED UP BETWEEN GRAPH VERSUS TABLE\]

@fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty within Geomarkets over time, revealing shifting dynamics across Geomarkets similar to those observed in median household income. In 1980, poverty rates were comparably low in both Collin & Rockwall Counties and Dallas County, moderate in the Irving, Arlington, and Grand Prarie and West of Dallas/Fort Worth Metroplex Geomarkets, and highest for City of Dallas and City of Fort Worth Geomarkets. Overtime, poverty rates declined in Collin & Rockwall Counties and West of Dallas/Ft. Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in a widening gap in poverty levels across the Dallas-Fort Worth metropolitan area. A similar widening disparity is evident evident in educational attainment. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the Collin & Rockwall Counties Geomarket experienced a substantially greater increase - from 23% in 1980 to 50% in 2020- maintaining the gap with other Geomarkets in terms of educational attainment.

We also analyze student list orders collected through public requests to assess how the racial/ethnic and socioeconomic composition of included versus excluded prospects varies when filtering on particular Geomarkets within the Dallas-Fort Worth metropolitan area. Specifically, we analyze three list orders placed by a research university targeting prospects across 87 different Geomarkets across the country, including all six Geomarkets in the Dallas-Fort Worth Metropolitan area. Each order also filtered for prospects in the 2019-2020 graduating class and based on PSAT scores across three ranges: 1070-1180, 1190-1260, and 1270-1520. Similar to Chicago analyses, we focus on prospects within the middle PSAT score range (1190-1260) from the Dallas-Fort Worth metropolitan area to simulate Geomarket contributions and assess patterns of inclusion versus exclusion.
